Declarative Pearl: Lightweight Probabilistic Programming in Curry

### 2015-08-14 Paper/paper.tex
Start with section on the idea of PFLP

The definition for the probabilistic functions \code{earthquake},
\code{burglary} and \code{alarm} follow straight-forward from the
scheme of the given model. %
The function \code{bernoulli} yields a probabilistic value that is
\code{True} with probability given by the first argument and
\code{False} with probability $1 - prob$. %

@@

@@

\begin{itemize}
\item we model probabilistic values with nondeterminstic values -- a
  probabilistic value is a pair of a value and its probability; a
  probabilistic value %

\begin{verbatim}
biasedCoin = (True, 0.3)
biasedCoin = (False, 0.7)
\end{verbatim}

  Rules with overlapping patterns on the left-hand side are one way to
  define nondeterministic functions in Curry. %
  Alternatively, Curry provides an explicite choice operator \code{?}
  that chooses nondeterministic between its first and second
  argument. %

\begin{verbatim}
coin = (True ? False,0.5)
\end{verbatim}

\item Working within a host language enables us to focus on the
  problem itself and frees us from reimplementing standard primitives
  (IO, data structures), essential design decisions (syntax,
  parser). %
  Based on which criteria do you choose the host language? %
  As for probabilistic languages that started growing over time,
  libraries that enable probablistic programming exist in the whole
  variety of paradigms: declarative as well as imperative. %
  For a start, you obviously choose the paradigm that has the clearest
  structure and enables straight-forward programming. %
  We want to address a wide range of developers and even
  non-PL-experts, thus, we choose a paradigm than can be learnt
  easily. %
  Thus, obvioulsy, you choose a declarative programming language. %
  By choosing Curry, we argue that it is sufficient to reuse built-in
  features in order to cover the essence of probabilistic programs,
  especially its semantic. %

@@

  The definition of \code{coin'} consists of an expression with a
  nondeterminstic choice on top-level position, whereas the definition
  of \cote{coin} generalises the common parts of the two possible
  values and uses the nondeterminism in a deeper position. %

>>>>

The definition for the probabilistic functions \code{earthquake},
\code{burglary} and \code{alarm} follow straight-forward from the
scheme of the given model. %
The function \code{bernoulli} yields a probabilistic value that is
\code{True} with probability given by the first argument and
\code{False} with the converse probability. %

@@

After defining the model, we want to ask queries about certian
propoerties and probabilities, respectively. %
The essence of probabilistic programming is to harmonise the problem
and its analysis: we want to define both, the problem and the queries,
in same language. %

@@

In the style of \cite{pfp}, the library's heart is a data type that
represents distributions: such a distribution consists of pair of
value and its probability. %

\begin{verbatim}
data Probability = Prop Float
data Dist a = Dist (a,Probability)
\end{verbatim}

However, the attentive reader may notice a slight difference: the PFLP
approach does not need to represent the collection of possible values
and their corresponding probabilities as list. %
In functional languages like Curry, the representation of several
results is commonly represented with nondeterminism instead of
directly using a container like lists. %
That is, we model probabilistic values as nondeterministic pairs of
possible value tagged with their probability. %

For example, we can define a biased coin with a $30\%-70\%$ chance to
land on head and tail, respectively, as follows. %

\begin{verbatim}
data Coin = Head | Tail
biasedCoin = (Head, 0.3)
biasedCoin = (Tail, 0.7)
\end{verbatim}

Rules with overlapping patterns on the left-hand side are one way to
define nondeterministic functions in Curry. %
Alternatively, Curry provides an explicite choice operator \code{?}
that chooses nondeterministicly between its first and second
argument. %
Hence, we can a define a fair coin with the following rule. %

\begin{verbatim}
fairCoin = (Head ? Tail,0.5)
\end{verbatim}

\begin{itemize}
\item Working within a host language enables us to focus on the
  problem itself and frees us from reimplementing standard primitives
  (IO, data structures), essential design decisions (syntax,
  parser). %
  Based on which criteria do you choose the host language? %
  As for probabilistic languages that started growing over time,
  libraries that enable probablistic programming exist in the whole
  variety of paradigms: declarative as well as imperative. %
  For a start, you obviously choose the paradigm that has the clearest
  structure and enables straight-forward programming. %
  We want to address a wide range of developers and even
  non-PL-experts, thus, we choose a paradigm than can be learnt
  easily. %
  Thus, obvioulsy, you choose a declarative programming language. %
  By choosing Curry, we argue that it is sufficient to reuse built-in
  features in order to cover the essence of probabilistic programs,
  especially its semantic. %

@@

  The definition of \code{coin'} consists of an expression with a
  nondeterminstic choice on top-level position, whereas the definition
  of \code{coin} generalises the common parts of the two possible
  values and uses the nondeterminism in a deeper position. %
